Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2019-03-21 versions/response.v2.txt

> Comments to the Author
> As with the previous version, I believe the topic matter will be of
> interest. For me, however, the work is incomplete in important ways
> that I believe can be relatively easily addressed. The authors went
> beyond my expectations in calculating explicit solutions for the
> limiting edge-lengths (Fig S2) when tau1 is the putative topology of
> estimation. This allows them to draw conclusions about edge-length
> inconsistency but there is no definitive conclusion about `strict'
> topological inconsistency (in the sense of converging upon the wrong
> tree rather than a multifurcating one). From the statements near l238
> it seems the authors have completed some preliminary investigation on
> strict topological inconsistency and find no evidence for it at
> present. Assuming that remains true, the findings of the current ms
> are of a different nature than the previous version where strict
> topological inconsistency was found. In settings with a limiting
> multifurcating topology, because it is multifurcating, for finite but
> large sequence-lengths, measures of uncertainty (eg. bootstrap
> support) will suggest that there is substantial uncertainty in
> whatever topology gets estimated (this probably deserves discussion).
> The joint optimization approach thus lacks the power to estimate the
> correct topology where ML would but is not being positively mislead.
> Assuming the lack of strict topological inconsistency holds, the
> results for the method are not as negative as they were in the
> previous version.

%EM I honestly can't figure out what the reviewer wants to see re the bootstrap. Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.

The reviewer is correct that this story is different than the one in the previous version of the paper.

%EM Shall we strike the speculation about higher likelihoods on the wrong topology?

@@

> -l150: I initially found this paragraph confusing. Table S1 is a
> result that you are using as intuition, so you can't be giving
> intuition for the entire result (otherwise it gets circular). I think
> you are trying to give intuition for the `everywhere except a set of
> measure' portion of the result rather than entire result. Perhaps `the
> proof that hat t neq t* except on a set of measure 0 can be seen
> intuitively...'.

We have added the suggested wording, thanks.

> -l154 and (should check elsewhere): `the limiting value of the
> estimator' rather than `the estimator' is what is being referred to
> here.

We focus in this paper in the limiting case only.
We have added a sentence clarifying language: (INSERT)

@@

> -l189: `depends on the values of other branch parameters' or some
> other rewording. I don't think it is that they are unknown. It is that
> they need optimization, so you can't optimize w independently of them.

We have adopted the suggested wording, thanks.

> -l201, Figs 2 and 3: I don't understand why direct numerical
> optimization was needed. You have the exact expressions for hat w in
> Fig S2. So you could have plotted them as Fig 2. A similar comment
> applies to Fig 3. Are you trying to convince the reader that your
> proofs have no errors? That seems more an issue for supp material.

We have rewritten this section to clarify what is being used where.
%EM das-- is this right?
Specifically, although we can find the regions for which the various ancestral state splits are optimal, we cannot obtain optimal branch lengths analytically.

We moved these figures around in response to a previous reviewer comment, who thought that Figure S2 was quite uninteresting while Figure 2 was relevant.
We would prefer to leave them be.
Figure 3 is an essential part of our argument that the internal branch length has substantial deviation from truth.

@@

> Comments to the Author
> Shaw et al provided a revised version of the manuscript that I
> reviewed about a year ago showing that joint estimation of ancestral
> states and branch length is in general not consistent. The authors now
> map out the regions in which topologies are incorrectly estimated and
> include a plot that shows the bias of the internal branch length for
> different tree configurations. I think this is a useful piece of work.
> But some parts are completely unsurprising, while some empirical
> results seem inaccurate or misleading.

> The most important results/claims by Shaw et al are:

> 1) "jointly estimating phylogenetic parameters along with ancestral
> states is not consistent in general"... "only correct on a set of
> measure 0".
> It is clearly true that branch length will be biased. But the fact
> that joint estimation will not recover the true branch lengths is not
> surprising but perfectly expected. Joint estimation ignores subleading
> configurations of internal states. Since the subleading terms
> typically require more state changes, branch length estimates ignoring
> them are biased. In a similar way a Taylor expansion is almost nowhere
> "correct", but still a useful and insightful approximation (a better
> analogy is ignoring looped diagram in a diagrammatic expansion.).

I'm afraid that this fact wasn't obvious to us at the outset of this project. The interest that the other reviewers have shown in the manuscript confirms that it is not broadly obvious.

@@

We have worked this out with Dr. Neher and found that the difference lay in that he was not optimizing the objective function from both sides.
This also demonstrated that the above-listed intuitive explanation was incorrect.

> 3) "[The bias is] of the same order as the branch lengths themselves."
> This is incorrect or misleading, depending on how you read it. In
> typically cases, the error made by this procedure should be quadratic
> in branch length and hence of higher order than the branch length
> itself. All empirical test we have done via numerically optimizing
> branch lengths suggests this is true for typical trees and that the
> errors made are small provided ALL adjacent branch lengths are short.
> That said, relative error of one branch can be large if adjacant
> branches are long such as the case above. In this case the error is
> quadratic in the length of the longer adjacent branches rather than
> the branch itself.

We have removed the "order" language, instead just focusing on the downward bias.

### 2019-03-21 joint_inf.tex
more progress through responses.

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to provably consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
More generally, we show that this joint method only estimates branch lengths correctly on a set of measure zero.
We show empirically that branch length estimates are biased even for short branches, with bias of the same order as the branch lengths themselves.
\end{abstract}

@@

An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters are the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be exactly zero.
We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero (i.e.\ a set that occupies zero volume in parameter space).
Empirically, we find areas where joint inference consistently underestimates interior branch lengths, including regions of short branch length where bias is on the same order as the branch length.

@@

@@

This result implies an inconsistency because the joint estimate of the interior branch length is zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

We provide the following as an intuition for the theoretical development.
For a particular site pattern, to obtain the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of the terms of the joint likelihood expression depending on which ancestral state corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the other branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.

\subsubsection*{Empirical validation}

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the analytically\hyp{}derived region (Fig.~\ref{fig:max-anc-state}).
To verify the regions of inconsistency and obtain a clearer picture of the closed form parameter estimates, we plot the optimal $\hat{w}$ via joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
As before, the region of inconsistency encompasses almost half of the branch fidelity space; given the correct topology, there are many situations where we estimate the interior branch length to be zero.

@@

In our optimization procedure, we again consider the 81 separate cases (Table~\ref{tab:likelihoods}) and, for each function, we compute the closed form solution for $\hat{t}$.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are on the same order as the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

In contrast, inference on the integrated likelihood performs as expected, such that $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
The errors in this case \eqref{eq:marginal_likelihood} via optimization with L-BFGS-B are lower than machine tolerance.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, we have shown that the only generating parameters that yield consistent branch length estimates given the correct topology lie in a set of measure zero.
In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in topological ambiguity: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when one set of sister branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Initial results by of Dr\. Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

>>>>

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to provably consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
More generally, we show that this joint method only estimates branch lengths correctly on a set of measure zero.
We show empirically that branch length estimates are systematically biased downward, even for short branches.
\end{abstract}

@@

An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters are the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
One would like to know when this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be exactly zero.
We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero (i.e.\ a set that occupies zero volume in parameter space).
Empirically, we find that joint inference consistently underestimates interior branch lengths, even when branch lengths are small.

@@

Because we are interested in the properties of joint inference in the limit of $n$ going to infinity, we will focus on estimates in this case and simply refer to the right hand side of \eqref{eq:exp-log-lik} as ``the estimator.''

@@

This result implies an inconsistency because the joint estimate of the interior branch length is zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (Table in bottom panel of Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent; we elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

We provide the following as an intuition for the theoretical development.
For a particular site pattern, to obtain the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of the terms of the joint likelihood expression depending on which ancestral state corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the other branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
% If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.

\subsubsection*{The outcome of joint inference is divided into distinct regions}

One can also use the possible cases for the form of the joint likelihood (Table~\ref{tab:likelihoods}) to understand the outcome of joint inference for various regions in parameter space.
For example, one can plot the regions in which each of these ancestral state splits attain the maximum (Fig.~\ref{fig:max-anc-state}).
In order to understand the optimal internal branch length in each of these cases, we optimize branch lengths numerically to obtain $\hat{w}$ under joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
Using either technique, we see that the region of inconsistency encompasses almost half of the branch fidelity space: given the correct topology, there are many situations where the interior branch length is estimated to be zero.

@@

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are comparable to the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

In contrast, inference on the integrated likelihood performs as expected, such that $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
The errors in this case \eqref{eq:marginal_likelihood} via optimization with L-BFGS-B are lower than machine tolerance.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, we have shown that the only generating parameters that yield consistent branch length estimates given the correct topology lie in a set of measure zero.
In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in topological ambiguity: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when one set of sister branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is comparable to the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Initial simulation-based results by Dr\. Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
On the other hand, when the joint inference optimization is started with a small branch length, the region ending up with a zero branch length is much larger, even for 4-state DNA models.
In a more mathematical direction, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
